---
title: "`r report_title`"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
    df_print: kable
---

<!--
  cellboard reproducible QC report.

  This template is rendered by cellboard::render_qc_report(), which injects the
  following objects into the knit environment:
    sce, thresholds, report_title, sample_name,
    compute_qc(), apply_qc_filter(), summarise_qc(), qc_metric_cols(),
    qc_violin(), qc_scatter(), qc_knee(), fmt_int()
  Re-running render_qc_report() with the same inputs reproduces this document.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", dpi = 96
)
suppressPackageStartupMessages({
  library(ggplot2)
  requireNamespace("SingleCellExperiment", quietly = TRUE)
  requireNamespace("SummarizedExperiment", quietly = TRUE)
})

# Make sure metrics exist; compute them if the caller passed a raw SCE.
cols <- qc_metric_cols()
have_metrics <- all(c(cols[["nUMI"]], cols[["nGene"]], cols[["pct_mito"]]) %in%
                      colnames(SummarizedExperiment::colData(sce)))
if (!have_metrics) {
  sce <- compute_qc(sce)
}

res     <- apply_qc_filter(sce, thresholds)
summ    <- summarise_qc(res)
sce_ann <- res$sce
qc_meta <- S4Vectors::metadata(sce)$cellboard$qc
```

## Dataset overview {#overview}

**Sample:** `r sample_name`

```{r overview}
ov <- data.frame(
  Property = c("Genes (features)", "Cells (barcodes)", "Assays",
               "Mitochondrial genes matched", "Doublet detection (scDblFinder)"),
  Value = c(
    fmt_int(nrow(sce)),
    fmt_int(ncol(sce)),
    paste(SummarizedExperiment::assayNames(sce), collapse = ", "),
    if (is.null(qc_meta)) "-" else as.character(qc_meta$n_mito_genes),
    if (is.null(qc_meta)) "-" else if (isTRUE(qc_meta$ran_doublets)) "yes" else "no"
  ),
  check.names = FALSE
)
knitr::kable(ov, align = c("l", "r"))
```

## QC metric distributions {#distributions}

Dashed red lines mark the active thresholds. Points are coloured by whether the
cell is **retained** or **discarded** under the current thresholds.

```{r violins, fig.show = "hold", out.width = "48%", fig.height = 4}
for (m in c("pct_mito", "nGene", "nUMI", "doublet_score")) {
  print(qc_violin(sce_ann, m, thresholds = thresholds))
}
```

```{r scatter-knee, fig.show = "hold", out.width = "48%", fig.height = 4.2}
print(qc_scatter(sce_ann, "nUMI", "nGene", colour = "status", thresholds = thresholds))
print(qc_knee(sce_ann, thresholds = thresholds))
```

## Thresholds applied {#thresholds}

```{r thresholds-table}
th <- thresholds
rows <- list(
  list("Minimum genes / cell",        th$min_genes),
  list("Maximum genes / cell",        th$max_genes),
  list("Minimum UMIs / cell",         th$min_umi),
  list("Maximum UMIs / cell",         th$max_umi),
  list("Maximum mitochondrial %",     th$max_mito),
  list("Maximum doublet score",       th$max_doublet),
  list("Remove scDblFinder doublets", if (isTRUE(th$remove_doublet_class)) "yes" else "no")
)
tbl <- do.call(rbind, lapply(rows, function(r) {
  data.frame(Threshold = r[[1]],
             Value = if (is.null(r[[2]]) || length(r[[2]]) == 0) "(inactive)" else as.character(r[[2]]),
             stringsAsFactors = FALSE)
}))
knitr::kable(tbl, align = c("l", "r"))
```

## Filtering outcome {#outcome}

```{r totals}
knitr::kable(summ$totals, col.names = c("", "Cells"), align = c("l", "r"))
```

```{r retained-bar, fig.height = 1.6, fig.width = 7}
tot <- summ$totals$value
bar <- data.frame(
  status = factor(c("Retained", "Discarded"), levels = c("Discarded", "Retained")),
  cells  = c(tot[2], tot[3])
)
ggplot(bar, aes(x = "", y = cells, fill = status)) +
  geom_col(width = 0.5) +
  coord_flip() +
  scale_fill_manual(values = c(Retained = "#2C7FB8", Discarded = "#D95F02"), name = NULL) +
  labs(x = NULL, y = "Cells") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", panel.grid = element_blank())
```

**`r fmt_int(res$n_keep)`** of **`r fmt_int(res$n_total)`** cells
(`r summ$totals$value[4]`%) pass QC.

### Cells failing each filter {#by-reason}

A cell may fail more than one filter, so these counts can sum to more than the
number of discarded cells.

```{r by-reason}
if (nrow(summ$by_reason)) {
  knitr::kable(summ$by_reason[, c("description", "cells_failing")],
               col.names = c("Filter", "Cells failing"), align = c("l", "r"))
} else {
  cat("No active filters.\n")
}
```

## Reproduce this analysis {#reproduce}

Running the snippet below on the same input reproduces this report's filtering
with the exact thresholds used here:

```{r repro-code, results = "asis"}
mk <- function(name, val) {
  if (is.null(val) || length(val) == 0) return(NULL)
  sprintf("  %-20s = %s", name, format(val))
}
args <- c(
  mk("min_genes", th$min_genes),
  mk("max_genes", th$max_genes),
  mk("min_umi",   th$min_umi),
  mk("max_umi",   th$max_umi),
  mk("max_mito",  th$max_mito),
  mk("max_doublet", th$max_doublet),
  sprintf("  %-20s = %s", "remove_doublet_class", isTRUE(th$remove_doublet_class))
)
code <- paste0(
  "library(cellboard)\n",
  "sce <- read_input(\"<your-input>\")   # 10x dir, .h5, or .rds (SCE/Seurat)\n",
  "sce <- compute_qc(sce)                # mito %, nGene, nUMI, doublet score\n",
  "th  <- qc_thresholds(\n",
  paste(args, collapse = ",\n"), "\n)\n",
  "res <- apply_qc_filter(sce, th)\n",
  "saveRDS(res$filtered, \"filtered_sce.rds\")\n"
)
cat("```r\n", code, "```\n", sep = "")
```

## Session information {#session}

```{r session}
sessionInfo()
```
