---
title: "cellboard cookbook: QC a PBMC dataset from scratch"
author: "Amit Shenoy"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{cellboard cookbook: QC a PBMC dataset from scratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4, dpi = 96, message = FALSE, warning = FALSE
)
```

This cookbook walks through a complete single-cell QC session with **cellboard**:
load a public PBMC dataset, inspect its quality, choose thresholds, export a
filtered object, and render a reproducible report — first as code, then in the
point-and-click [dashboard](#dashboard). Two bonus sections show how to
[schedule the report on Posit Connect](#schedule) and how to
[slot cellboard into an existing Seurat pipeline](#seurat).

```{r library}
library(cellboard)
```

## 1. Get a real PBMC dataset {#data}

The canonical teaching dataset is 10x Genomics' **pbmc3k** (~2,700 peripheral
blood mononuclear cells). This vignette downloads the **real** data straight
from 10x and caches it under `tools::R_user_dir()`, falling back to the bundled
example only if the download is unavailable (e.g. a fully offline build):

```{r get-data}
get_pbmc3k <- function() {
  cache <- tools::R_user_dir("cellboard", "cache")
  dir.create(cache, recursive = TRUE, showWarnings = FALSE)
  mtx_parent <- file.path(cache, "filtered_gene_bc_matrices")
  if (!dir.exists(mtx_parent)) {
    tar <- file.path(cache, "pbmc3k.tar.gz")
    url <- paste0("https://cf.10xgenomics.com/samples/cell/pbmc3k/",
                  "pbmc3k_filtered_gene_bc_matrices.tar.gz")
    options(timeout = max(600, getOption("timeout")))
    utils::download.file(url, tar, mode = "wb", quiet = TRUE)
    utils::untar(tar, exdir = cache)
  }
  read_input(mtx_parent)   # read_input() descends into the nested hg19/ folder
}

sce <- tryCatch(get_pbmc3k(), error = function(e) {
  message("pbmc3k download unavailable (", conditionMessage(e),
          "); using the bundled example instead.")
  example_sce()
})
sce
```

```{r which-data, echo = FALSE, results = "asis"}
if (ncol(sce) > 1500) {
  cat(sprintf("> Loaded the **real 10x pbmc3k** dataset: %d genes x %d cells.\n",
              nrow(sce), ncol(sce)))
} else {
  cat(sprintf(paste0("> Download unavailable -- using the bundled simulated ",
                     "example (%d cells). Every step below is identical.\n"),
              ncol(sce)))
}
```

Other ways to obtain the same object:

```{r download-alt, eval = FALSE}
## From Bioconductor (downloads via ExperimentHub):
BiocManager::install("TENxPBMCData")
sce <- TENxPBMCData::TENxPBMCData("pbmc3k")

## You already have an .rds or a Seurat object:
sce <- read_input("my_sce.rds")
sce <- read_input("my_seurat.rds", type = "seurat")
```

`read_input()` always returns a `SingleCellExperiment` with a `counts` assay,
whether you start from a 10x directory (it descends into nested matrix folders
such as `hg19/`), a `.h5` file, an existing SCE, or a Seurat object.

## 2. Compute QC metrics {#compute}

`compute_qc()` adds four standardised per-cell metrics to `colData`:
`nUMI`, `nGene`, `pct_mito`, and an `scDblFinder` `doublet_score` /
`doublet_class`.

```{r compute}
sce <- compute_qc(sce)               # mito pattern defaults to ^MT-/^mt-

cd <- as.data.frame(SummarizedExperiment::colData(sce))
round(sapply(cd[, c("nUMI", "nGene", "pct_mito", "doublet_score")], summary), 2)
```

Mitochondrial genes are matched against gene **symbols** (from a `Symbol`-like
`rowData` column when present, else the row names), so this works whether your
features are Ensembl IDs or symbols. For mouse, pass `mito_pattern = "^mt-"`.

## 3. Inspect the distributions {#inspect}

The same plot helpers the dashboard uses are exported, so you can eyeball QC
before deciding on cutoffs:

```{r violins, fig.height = 3.4}
qc_violin(sce, "pct_mito")
qc_violin(sce, "nGene")
```

```{r knee, fig.height = 3.6}
qc_knee(sce)   # barcode-rank plot
```

In pbmc3k most cells are healthy (median ~2% mitochondrial reads); a tail sits
at high mitochondrial percentage (dying or lysed cells) and `scDblFinder` flags
a small set of likely doublets -- exactly what the thresholds below remove.

## 4. Choose thresholds {#thresholds}

`default_thresholds()` gives a conservative starting point (and can suggest an
upper UMI bound from the data). Here we use the cutoffs from the classic Seurat
pbmc3k tutorial -- 200-2,500 genes and < 5% mitochondrial reads -- and
additionally drop scDblFinder-called doublets. A cell is discarded if it fails
**any** active threshold.

```{r thresholds}
th <- qc_thresholds(
  min_genes            = 200,
  max_genes            = 2500,
  max_mito             = 5,
  remove_doublet_class = TRUE   # drop scDblFinder-called doublets
)

res <- apply_qc_filter(sce, th)
c(input = res$n_total, retained = res$n_keep, discarded = res$n_discard)
```

`summarise_qc()` breaks down *why* cells were dropped (a cell can fail more than
one filter):

```{r summarise}
s <- summarise_qc(res)
knitr::kable(s$by_reason)
```

Visualise the decision. Points are coloured retained vs discarded, with the
threshold lines overlaid:

```{r scatter, fig.height = 4.2}
qc_scatter(res$sce, "nUMI", "nGene", colour = "status", thresholds = th)
```

Tightening a threshold can only ever remove more cells — handy when calibrating:

```{r monotone}
strict <- apply_qc_filter(sce, qc_thresholds(min_genes = 500, max_mito = 5))
c(lenient = res$n_keep, strict = strict$n_keep)
```

## 5. Export the filtered object {#export}

```{r export, eval = FALSE}
filtered <- res$filtered             # a filtered SingleCellExperiment
saveRDS(filtered, "pbmc_filtered_sce.rds")

# per-cell metrics, e.g. for a supplementary table
write.csv(as.data.frame(SummarizedExperiment::colData(res$sce)),
          "pbmc_cell_metrics.csv")
```

```{r export-dims}
dim(res$filtered)
```

## 6. Render the reproducible report {#report}

`render_qc_report()` knits a self-contained HTML report documenting the metrics,
the thresholds, and the retained-vs-discarded outcome. Re-running it with the
same inputs reproduces the document byte-for-byte (modulo timestamps).

```{r report, eval = FALSE}
render_qc_report(
  res$sce, th,
  output_file = "pbmc_qc_report.html",
  sample_name = "pbmc3k"
)
```

## 7. ...or do all of this in the dashboard {#dashboard}

Every step above is available point-and-click:

```{r run-app, eval = FALSE}
run_app()
```

Click **Load example PBMC** (or upload your data), press **Compute QC metrics**,
then drag the threshold sliders on the **Filtering** tab and watch the plots,
KPI boxes and cell table update live. The **Export** tab downloads the filtered
object, the metrics CSV, and the HTML report.

## Bonus: schedule the report on Posit Connect {#schedule}

The report template is a parameterised R Markdown document, so Connect can run
it on a schedule.

1. **Publish the template.** From the RStudio IDE, open
   `inst/report/qc_report.Rmd` and click *Publish*, or:

    ```{r connect-publish, eval = FALSE}
    rsconnect::deployDoc("inst/report/qc_report.Rmd")
    ```

2. **Schedule it.** In Connect, open the published report →
   *Settings → Schedule → Run report on a schedule* (e.g. weekly), and add
   e-mail recipients under *Sharing*. Connect re-knits the report on the
   schedule and e-mails the refreshed HTML.

3. **Parameterise per run.** The report reads its data and thresholds from the
   knit environment; to point a scheduled run at fresh data, deploy a thin
   wrapper `.Rmd` that calls `cellboard::render_qc_report()` on a path your
   pipeline updates (e.g. the latest run's `.rds` on a shared mount).

The whole **dashboard** deploys the same way — see the README's
*Deploy to Posit Connect* section (`rsconnect::deployApp()` or the included
`deploy-connect` GitHub Action).

## Bonus: integrate with an existing Seurat pipeline {#seurat}

cellboard speaks `SingleCellExperiment`, but converts Seurat objects for you.
Drop it into a Seurat workflow to QC with `scDblFinder` and carry the decision
back:

```{r seurat, message = FALSE, warning = FALSE}
library(Seurat)

# Pretend this came from your pipeline (here: build a small one from the example)
counts      <- SummarizedExperiment::assay(example_sce(), "counts")
seurat_obj  <- SeuratObject::CreateSeuratObject(counts = counts)

# 1. hand the object to cellboard
sce <- as_sce(seurat_obj)            # Seurat -> SingleCellExperiment
sce <- compute_qc(sce)

# 2. decide which cells to keep
keep <- apply_qc_filter(sce, qc_thresholds(min_genes = 200, max_mito = 10))$keep

# 3. carry the decision back into the Seurat object
seurat_qc <- seurat_obj[, keep]
seurat_qc
```

You can also push cellboard's metrics back as Seurat metadata:

```{r seurat-meta}
md <- as.data.frame(SummarizedExperiment::colData(sce))[, c("nUMI", "nGene", "pct_mito")]
seurat_obj <- SeuratObject::AddMetaData(seurat_obj, md)
head(seurat_obj[[]], 3)
```

From here, continue your normal Seurat (or Bioconductor `scran`/`scater`)
analysis on the QC'd cells.

## Session information

```{r session}
sessionInfo()
```
